---
title: "Hierarchical Visualization with SunburstR"
author: "Julio Carrión"
date: "2025-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document demonstrates how to generate hierarchical sunburst visualizations from biodiversity data.
The dataset includes BIN identifiers, localities, biomes, ecoregions, and full taxonomy.

## Required Libraries

```{r}
library(dplyr)
library(tidyr)
library(plotly)
library(sunburstR)
```

## Load Data

```{r}
# Load the combined dataset
combined_data <- readRDS("combined_data.rds")  
# Expected columns: bin_uri, locality, biome, ecoregion, kingdom, phylum, class, order, family, genus, species
```

## Create Taxonomic Hierarchy

```{r}
# Create a column with the full taxonomic hierarchy
combined_data <- combined_data %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-"))
```

## Costa Rica - Cocos Island

```{R}
tabla_sunburst_IslaCocos <- combined_data %>%
  filter(locality == "IslaCocos") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_IslaCocos, count = TRUE)
```

## Costa Rica - Guancaste

```{R}
tabla_sunburst_Guanacaste <- combined_data %>%
  filter(locality == "Guanacaste") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Guanacaste, count = TRUE)
```




## Costa Rica - Quetzales

```{R}
tabla_sunburst_Quetzales <- combined_data %>%
  filter(locality == "Quetzales") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Quetzales, count = TRUE)
```


## Costa Rica - Baru

```{R}
tabla_sunburst_Baru <- combined_data %>%
  filter(locality == "Baru") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Baru, count = TRUE)
```

## Panama - Barro Colorado Island

```{R}
tabla_sunburst_BarroColorado <- combined_data %>%
  filter(locality == "BarroColorado") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_BarroColorado, count = TRUE)
```


## Ecuador - Galapagos

```{R}
tabla_sunburst_Galapagos <- combined_data %>%
  filter(locality == "Galapagos") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Galapagos, count = TRUE)
```


## Ecuador - CerroBlanco

```{R}
tabla_sunburst_CerroBlanco <- combined_data %>%
  filter(locality == "CerroBlanco") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_CerroBlanco, count = TRUE)
```


## Ecuador - Mashpi

```{R}
tabla_sunburst_Mashpi <- combined_data %>%
  filter(locality == "Mashpi") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Mashpi, count = TRUE)
```



## Ecuador - Pululahua

```{R}
tabla_sunburst_Pululahua <- combined_data %>%
  filter(locality == "Pululahua") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Pululahua, count = TRUE)
```



## Ecuador - Paluguillo

```{R}
tabla_sunburst_Paluguillo <- combined_data %>%
  filter(locality == "Paluguillo") %>%
  mutate(path = paste(kingdom, phylum, class, order, family, genus, species, sep = "-")) %>%
  group_by(path) %>%
  summarise(count = n(), .groups = "drop")

# Plot sunburst
sunburst(tabla_sunburst_Paluguillo, count = TRUE)
```








